Analysis: Machine Learning - OblAmt - Outliers

### 2021-12-12 code/analysis_code/analysis-ML-OA-outlier.Rmd
Last ML model update

* Null
* Decision Tree
* Random Forest
* Bagged Tree

It compares the  models, and then finally fits the “best” model to the test data.

@@

## Load Data
Load the analysis data from the `processed_data` folder in the project file.
```{r load data}
#path to data
#note the use of the here() package and not absolute paths
data_location <- here::here("data","processed_data","analysisdata.rds")

@@

* Drop the unused variables
* Perform feature engineering
* Transform Incident Month into ordinal variable
* Set the random seed to 123
* Split the dataset into 70% training, 30% testing, stratified on incidentType
* 5-fold cross validation, 5 times repeated, stratified on incidentType for the CV folds
* Create a recipe for data and fitting that codes categorical variables as dummy variables
```{r data setup}
#requested amount filter
obl_data <- analysisdata %>%
              dplyr::select(state, IncidentYear, IncidentMonth, declarationType, incidentType, IncidentDuration, ResponseDays, IH, PA, HM,
                            Population, Counties, TotalAgencies, FEMACSAvg, OblAmt)

#transform incident month into ordinal variable
obl_data$IncidentMonth <- ordered(obl_data$IncidentMonth, levels = 1:12,
                                  labels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"))

#not enough values per state, so create a variable that is FEMA Regions
obl_data$FEMARegion <- ifelse((obl_data$state == "CT" |
                                 obl_data$state == "ME" |
                                 obl_data$state == "MA" |
                                 obl_data$state == "NH" |
                                 obl_data$state == "RI" |
                                 obl_data$state == "VT"),
                              "Region I",
                              ifelse((obl_data$state == "NJ" |
                                         obl_data$state == "NY" |
                                         obl_data$state == "PR" |
                                         obl_data$state == "VI"),
                                      "Region II",
                                       ifelse((obl_data$state == "DE" |
                                                 obl_data$state == "MD" |
                                                 obl_data$state == "PA" |
                                                 obl_data$state == "VA" |
                                                 obl_data$state == "DC" |
                                                 obl_data$state == "WV"),
                                              "Region III",
                                               ifelse((obl_data$state == "AL" |
                                                         obl_data$state == "FL" |
                                                         obl_data$state == "GA" |
                                                         obl_data$state == "KY" |
                                                         obl_data$state == "MS" |
                                                         obl_data$state == "NC" |
                                                         obl_data$state == "SC" |
                                                         obl_data$state == "TN"),
                                                      "Region IV",
                                                      ifelse((obl_data$state == "IL" |
                                                               obl_data$state == "IN" |
                                                               obl_data$state == "MI" |
                                                               obl_data$state == "MN" |
                                                               obl_data$state == "OH" |
                                                               obl_data$state == "WI"),
                                                            "Region V",
                                                             ifelse((obl_data$state == "AR" |
                                                                         obl_data$state == "LA" |
                                                                         obl_data$state == "NM" |
                                                                         obl_data$state == "OK" |
                                                                         obl_data$state == "TX"),
                                                                      "Region VI",
                                                                       ifelse((obl_data$state == "IA" |
                                                                                 obl_data$state == "KS" |
                                                                                 obl_data$state == "MO" |
                                                                                 obl_data$state == "NE"),
                                                                              "Region VII",
                                                                               ifelse((obl_data$state == "CO" |
                                                                                         obl_data$state == "MT" |
                                                                                         obl_data$state == "ND" |
                                                                                         obl_data$state == "SD" |
                                                                                         obl_data$state == "UT" |
                                                                                         obl_data$state == "WY"),
                                                                                      "Region VIII",
                                                                                       ifelse((obl_data$state == "AZ" |
                                                                                                 obl_data$state == "CA" |
                                                                                                 obl_data$state == "HI" |
                                                                                                 obl_data$state == "NV"),
                                                                                              "Region IX",
                                                                                               ifelse((obl_data$state == "AK" |
                                                                                                         obl_data$state == "ID" |
                                                                                                         obl_data$state == "OR" |
                                                                                                         obl_data$state == "WA"),
                                                                                                      "Region X",
                                                                                                      NA
                                                                                                      )
                                                                                              )
                                                                                      )
                                                                              )
                                                                      )
                                                            )
                                                      )
                                              )
                                     )
                              )

#drop state variable
obl_data <- obl_data %>%
              dplyr::select(-c(state))

#add to "other" category for incident Types that are infrequent
obl_data$incidentType[(obl_data$incidentType == "Chemical" |
                        obl_data$incidentType == "Coastal Storm" |
                        obl_data$incidentType == "Dam/Levee Break" |
                        obl_data$incidentType == "Mud/Landslide" |
                        obl_data$incidentType == "Volcano" |
                        obl_data$incidentType == "Terrorist" |
                        obl_data$incidentType == "Snow")] <- "Other"

@@

#split dataset into 66.6% training, 33.3% testing
#use incidentType as stratification
obl_data_split <- rsample::initial_split(obl_data, prop = 2/3,
                                     strata = incidentType)

@@

#5-fold cross validation, repeated 5 times, stratified by incidentType
obl_folds <- rsample::vfold_cv(obl_train_data,
                               strata = incidentType,
                               v = 5,
                               repeats = 5)

#create recipe that codes categorical variables as dummy variables
OA_rec <- recipes::recipe(OblAmt ~ ., data = obl_train_data) %>%
          recipes::step_dummy(all_nominal_predictors())
```

@@

#RMSE for training data for formatting later
obl_null_RMSE_train <- tibble::tibble(
                          rmse = rmse_vec(truth = obl_train_data$OblAmt,
                                              estimate = rep(mean(obl_train_data$OblAmt), nrow(obl_train_data))),
                          SE = 0,
                          model = "Null - Train")

@@

#RMSE for testing data for formatting later
obl_null_RMSE_test <- tibble::tibble(
                          rmse = rmse_vec(truth = obl_test_data$OblAmt,
                                              estimate = rep(mean(obl_test_data$OblAmt), nrow(obl_test_data))),
                          SE = 0,
                          model = "Null - Test")
```

@@

#define the tree model
tree_mod <-
  parsnip::decision_tree(cost_complexity = tune(),
                         tree_depth = tune(),
                         min_n = tune()) %>%
  parsnip::set_engine("rpart") %>%
  parsnip::set_mode("regression")

#use the recipe specified earlier (line 133)
```
---

@@

### 4. Tuning Using Cross-Validation and the `tune_grid()` function
```{r}
#tune the model with previously specified cross-validation and RMSE as target metric
tree_res <- tree_wflow %>%
                tune::tune_grid(resamples = obl_folds,
                                grid = tree_grid,
                                control = control_grid(verbose = TRUE),
                                metrics = yardstick::metric_set(rmse))

@@

#save file
DT_auto_fig_file = here("results","DT-auto-plot-obl.png")
ggsave(filename = DT_auto_fig_file, plot = DT_auto_plot)

#more detailed plot
DT_detail_plot <- tree_res %>%
                    workflowsets::collect_metrics() %>%
                    dplyr::mutate(tree_depth = factor(tree_depth)) %>%
                    ggplot2::ggplot(aes(cost_complexity, mean, color = tree_depth)) +
                             geom_line(size = 1.5, alpha = 0.6) +
                             geom_point(size = 2) +
                             facet_wrap(~ .metric, scales = "free", nrow = 2) +
                             scale_x_log10(labels = scales::label_number()) +
                             scale_color_viridis_d(option = "plasma", begin = 0.9, end = 0)
DT_detail_plot

#save file
DT_detail_fig_file = here("results","DT-det-plot-obl.png")
ggsave(filename = DT_detail_fig_file, plot = DT_detail_plot)

@@

### 6. Model evaluation
```{r}
#plot the tree
rpart.plot::rpart.plot(x = workflowsets::extract_fit_parsnip(best_tree_fit)$fit,
                                         roundint = F,
                                         type = 5,
                                         digits = 5,
                                         main = "Selected Decision Tree Model")
#save file manually -- ggsave throws an error

#find predictions and intervals
tree_resid <- best_tree_fit %>%
                  broom.mixed::augment(new_data = obl_train_data) %>%
                  dplyr::select(.pred, OblAmt) %>%
                  dplyr::mutate(.resid = .pred - OblAmt)

#plot model predictions from tuned model versus actual outcomes
#geom_abline draws a 45 degree line, along which the results should fall
DT_pred_act <- ggplot2::ggplot(tree_resid, aes(x = .pred, y = OblAmt)) +
                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
                        geom_point() +
                        labs(title = "Decision Tree Fit: Predicted vs. Actual Obligated Funds",
                              x = "Predicted Obligated Funds ($)",
                              y = "Actual Obligated Funds ($)")
DT_pred_act

#save file
DT_pred_act_file = here("results","DT-pred-act-obl.png")
ggsave(filename = DT_pred_act_file, plot = DT_pred_act)

#plot model with residuals
#the geom_hline plots a straight horizontal line along which the results should fall
DT_resid <- ggplot2::ggplot(tree_resid, aes(x = as.numeric(row.names(tree_resid)), y = .resid))+
                     geom_hline(yintercept = 0, color = "red", lty = 2) +
                     geom_point() +
                     labs(title = "Decision Tree Fit: Residuals",
                          x = "Observation Number",
                          y = "Residual")
DT_resid

#save file
DT_resid_file = here("results","DT-resid-obl.png")
ggsave(filename = DT_resid_file, plot = DT_resid)

#plot model fit vs residuals
#the geom_hline plots a straight horizontal line along which the results fall
DT_pred_resid <- ggplot2::ggplot(tree_resid, aes(x = .pred, y = .resid))+
                          geom_hline(yintercept = 0, color = "red", lty = 2) +
                          geom_point() +
                          labs(title = "Decision Tree Fit: Residuals vs Fitted Obligated Funds",
                                x = "Predicted Obligated Funds ($)",
                                y = "Residual")
DT_pred_resid

#save file
DT_pred_resid_file = here("results","DT-pred-resid-obl.png")
ggsave(filename = DT_pred_resid_file, plot = DT_pred_resid)

@@

@@

### 4. Tuning Using Cross-Validation and the `tune_grid()` function
```{r}
#tune the model with previously specified cross-validation and RMSE as target metric
RF_res <- RF_wflow %>%
              tune::tune_grid(resamples = obl_folds,
                              grid = RF_grid,
                              control = control_grid(verbose = TRUE, save_pred = TRUE),
                              metrics = metric_set(rmse))

@@

#save file
RF_auto_fig_file = here("results","RF-auto-plot-obl.png")
ggsave(filename = RF_auto_fig_file, plot = RF_auto_plot)
```

@@

#plot most important predictors in the model
RF_plot <- vip::vip(x_RF, num_features = 20)
RF_plot

#save  manually - ggsave throws an error

#find predictions and intervals
RF_resid <- best_RF_fit %>%
                broom.mixed::augment(new_data = obl_train_data) %>%
                dplyr::select(.pred, OblAmt) %>%
                dplyr::mutate(.resid = .pred - OblAmt)

#plot model predictions from tuned model versus actual outcomes
#geom_abline is a 45 degree line, along which the results should fall
RF_pred_act <- ggplot2::ggplot(RF_resid, aes(x = .pred, y = OblAmt)) +
                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) +
                        geom_point() +
                        labs(title = "RF Fit: Actual vs. Predicted Obligated Funds",
                              x = "Predicted Obligated Funds ($)",
                              y = "Actual Obligated Funds ($)")

#save file
RF_pred_act_file = here("results","RF-pred-act-obl.png")
ggsave(filename = RF_pred_act_file, plot = RF_pred_act)

#plot model with residuals
#the geom_hline plots a straight horizontal line along which the results should fall
RF_resids <- ggplot2::ggplot(RF_resid, aes(x = as.numeric(row.names(RF_resid)), y = .resid))+
                            geom_hline(yintercept = 0, color = "red", lty = 2) +
                            geom_point() +
                            labs(title = "RF Fit: Residuals",
                                  x = "Observation Number",
                                  y = "Residual")
RF_resids

#save file
RF_resid_file = here("results","RF-resid-obl.png")
ggsave(filename = RF_resid_file, plot = RF_resids)

#plot model fit vs residuals
#the geom_hline plots a straight horizontal line along which the results fall
RF_pred_resid <- ggplot2::ggplot(RF_resid, aes(x = .pred, y = .resid))+
                          geom_hline(yintercept = 0, color = "red", lty = 2) +
                          geom_point() +
                          labs(title = "Random Forest Fit: Residuals vs Fitted Obligated Funds",
                                x = "Fitted Requested Funds ($)",
                                y = "Residual")
RF_pred_resid

#save file
RF_pred_resid_file = here("results","RF-pred-resid-obl.png")
ggsave(filename = RF_pred_resid_file, plot = RF_pred_resid)

@@

@@

#start over with data split to allow for bootstrapping
#set random seed
set.seed(20140102)

#use bootstrap instead of CV
obl_data_bootsplit <- obl_train_data %>%
                        rsample::bootstraps(times = 10,
                                            strata = "incidentType")

#define model
bag_tree_mod <- baguette::bag_tree(cost_complexity = tune(),
                                   tree_depth = tune(),
                                   min_n = tune()) %>%
                parsnip::set_engine("rpart",
                                    times = 10) %>%
                parsnip::set_mode("regression")

#create recipe that codes categorical variables as dummy variables
BT_rec <- recipes::recipe(OblAmt ~ ., data = obl_train_data) %>%
          step_string2factor(all_nominal())
```
---

### 2. Workflow Definition
```{r}
#define workflow for tree
BT_wflow <- workflows::workflow() %>%
               workflows::add_model(bag_tree_mod) %>%
               workflows::add_recipe(BT_rec)
```

@@

### 3. Tuning Grid Specification
```{r}
#tuning grid specification
tree_grid <- dials::grid_regular(cost_complexity(),
                                 tree_depth(),
                                 min_n(),
                                 levels = 5)
```

@@

### 4. Tuning Using Cross-Validation and the `tune_grid()` function
```{r}
#tune the model with previously specified cross-validation and RMSE as target metric
BT_res <- baguette::bag_tree(cost_complexity = tune(),
                                   tree_depth = tune(),
                                   min_n = tune()) %>%
          parsnip::set_engine("rpart",
                              times = 10) %>%
          parsnip::set_mode("regression") %>%
          tune::tune_grid(preprocessor = BT_rec,
                          resamples = obl_data_bootsplit,
                          grid = tree_grid,
                          metrics = metric_set(rmse))

#collect metrics
BT_res %>% workflowsets::collect_metrics()

#default visualization
BT_auto_plot <- BT_res %>% autoplot()
BT_auto_plot

#save file
BT_auto_fig_file = here("results","BT-auto-plot-obl.png")
ggsave(filename = BT_auto_fig_file, plot = BT_auto_plot)

#more detailed plot
BT_det_plot <- BT_res %>%
                    workflowsets::collect_metrics() %>%
                    dplyr::mutate(tree_depth = factor(tree_depth)) %>%
                    ggplot2::ggplot(aes(cost_complexity, mean, color = tree_depth)) +
                             geom_line(size = 1.5, alpha = 0.6) +
                             geom_point(size = 2) +
                             facet_wrap(~ .metric, scales = "free", nrow = 2) +
                             scale_x_log10(labels = scales::label_number()) +
                             scale_color_viridis_d(option = "plasma", begin = 0.9, end = 0)
BT_det_plot

#save file
BT_det_fig_file = here("results","BT-det-plot-obl.png")
ggsave(filename = BT_det_fig_file, plot = BT_det_plot)
```

@@

### 5. Identify Best Model
```{r}
#select the BT model with the lowest rmse
BT_lowest_rmse <- BT_res %>%
                        tune::select_best("rmse")

#finalize the workflow by using the selected BT model
best_BT_wflow <- BT_wflow %>%
                      tune::finalize_workflow(BT_lowest_rmse)
best_BT_wflow

#one last fit on the training data
best_BT_fit <- best_BT_wflow %>%
                    parsnip::fit(data = obl_train_data)
```

@@

### 6. Model evaluation
```{r}
#extract model from final fit
x_BT <- best_BT_fit$fit$fit$fit

#vip package doesn't support BT

#find predictions and intervals
BT_resid <- best_BT_fit %>%
                  broom.mixed::augment(new_data = obl_train_data) %>%
                  dplyr::select(.pred, OblAmt) %>%
                  dplyr::mutate(.resid = .pred - OblAmt)

#plot model predictions from tuned model versus actual outcomes
#geom_abline draws a 45 degree line, along which the results should fall
BT_pred_act <- ggplot2::ggplot(BT_resid, aes(x = .pred, y = OblAmt)) +
                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
                        geom_point() +
                        labs(title = "Bagged Tree Fit: Predicted vs. Actual Obligated Funds",
                              x = "Predicted Obligated Funds ($)",
                              y = "Actual Obligated Funds ($)")
BT_pred_act

#save file
BT_pred_act_file = here("results","BT-pred-act-obl.png")
ggsave(filename = BT_pred_act_file, plot = BT_pred_act)

#plot model with residuals
#the geom_hline plots a straight horizontal line along which the results should fall
BT_resids <- ggplot2::ggplot(BT_resid, aes(x = as.numeric(row.names(BT_resid)), y = .resid))+
                      geom_hline(yintercept = 0, color = "red", lty = 2) +
                      geom_point() +
                      labs(title = "Bagged Tree Fit: Residuals",
                            x = "Observation Number",
                            y = "Residual")
BT_resids

#save file
BT_resid_file = here("results","BT-resids-obl.png")
ggsave(filename = BT_resid_file, plot = BT_resids)

#plot model fit vs residuals
#the geom_hline plots a straight horizontal line along which the results fall
BT_pred_resid <- ggplot2::ggplot(BT_resid, aes(x = .pred, y = .resid))+
                          geom_hline(yintercept = 0, color = "red", lty = 2) +
                          geom_point() +
                          labs(title = "Bagged Tree Fit: Residuals vs Fitted Obligated Funds",
                                x = "Predicted Obligated Funds ($)",
                                y = "Residual")
BT_pred_resid

#save file
BT_pred_resid_file = here("results","BT-pred-resid-obl.png")
ggsave(filename = BT_pred_resid_file, plot = BT_pred_resid)

#print model performance
#print 10 best performing hyperparameter sets
BT_res %>%
  tune::show_best(n = 10) %>%
  dplyr::select(rmse = mean, std_err, cost_complexity) %>%
  dplyr::mutate(rmse = round(rmse, 3),
                std_err = round(std_err, 4),
                cost_complexity = scales::scientific(cost_complexity))

#print the best model performance
BT_performance <- BT_res %>% tune::show_best(n = 1)
print(BT_performance)

#compare model performance to null model
BT_RMSE <- BT_res %>%
                tune::show_best(n = 1) %>%
                dplyr::transmute(
                  rmse = round(mean, 3),
                  SE = round(std_err, 4),
                  model = "Bagged Tree") %>%
               dplyr::bind_rows(RF_RMSE)
BT_RMSE

#save combined table
model_comp_file = here("results", "ML-model-comp-obl.Rds")
saveRDS(BT_RMSE, file = model_comp_file)
```

In examining the models, the Bagged Tree has the best RMSE, but it also has a higher SE than the other models. We will still choose the bagged tree as our model on which to evaluate the test data.

--

## Model Selection and Evaluation
Using the bagged tree model
```{r}
#fit model to training set but evaluate on the test data
BT_fit_test <- best_BT_wflow %>%
                    tune::last_fit(split = obl_data_split)

#compare test performance against training performance
BT_rmse_test <- collect_metrics(BT_fit_test) %>%
  dplyr::select(rmse = .estimate) %>%
  dplyr::mutate(data = "test")

>>>>

* Null
* Decision Tree
* Random Forest
* Elastic Net
* LASSO

It compares the  models, and then finally fits the “best” model to the test data. It also removes the clear outliers in the original analysis where the log(OblAmt) was less than zero.

@@

## Load Data
Load the OblAmt analysis data from the `processed_data` folder in the project file.
```{r load data}
#path to data
#note the use of the here() package and not absolute paths
data_location <- here::here("data","processed_data","analysisdata-OA.rds")

@@

* Log transform the outcome
* Drop `OblAmt`, `disasterNumber`, `declarationTitle`
* Remove the obvious outliers (-log(OblAmt))
* Set the random seed to 123
* Split the dataset into 70% training, 30% testing
* 5-fold cross validation for the spatial CV folds
* Create a recipe for data and fitting that codes categorical variables as dummy variables
```{r data setup}
#log transform OblAmt
analysisdata$logOblAmt <- log(analysisdata$OblAmt)

#drop unnecessary variables
obl_data_2 <- analysisdata %>%
              dplyr::select(-c(OblAmt, disasterNumber, declarationTitle))

#specify all character variables as factors and clean variable names
obl_data_1 <- obl_data_2 %>% 
                janitor::clean_names() %>%
                dplyr::mutate_if(is.character, factor)

#remove outliers (rows that have a -log(oblamt))
obl_data <- obl_data_1 %>%
              dplyr::filter(log_obl_amt > 0)

@@

#split dataset into 66.6% training, 33.3% testing
#use incidentType as stratification
obl_data_split <- rsample::initial_split(obl_data, prop = 2/3)

@@

#5-fold spatial cross validation
obl_folds_spatial <- spatialsample::spatial_clustering_cv(obl_train_data,
                                                          coords = c(latitude, longitude),
                                                          v = 5)

#create recipe that codes categorical variables as dummy variables
OA_rec <- recipes::recipe(log_obl_amt ~ ., data = obl_train_data) %>%
          recipes::step_unknown(state, new_level = "unknown state") %>%
          recipes::step_dummy(all_nominal_predictors())
```

@@

#RMSE for training data for formatting later
obl_null_RMSE_train <- tibble::tibble(
                          rmse = rmse_vec(truth = obl_train_data$log_obl_amt,
                                              estimate = rep(mean(obl_train_data$log_obl_amt), nrow(obl_train_data))),
                          SE = 0,
                          model = "Null - Train")

@@

#RMSE for testing data for formatting later
obl_null_RMSE_test <- tibble::tibble(
                          rmse = rmse_vec(truth = obl_test_data$log_obl_amt,
                                              estimate = rep(mean(obl_test_data$log_obl_amt), nrow(obl_test_data))),
                          SE = 0,
                          model = "Null - Test")
```

@@

#define the tree model
tree_mod <-
  parsnip::decision_tree(cost_complexity = tune(),
                         tree_depth = tune(),
                         min_n = tune()) %>%
  parsnip::set_engine("rpart") %>%
  parsnip::set_mode("regression")
```

@@

### 4. Tuning Using Cross-Validation and the `tune_grid()` function
```{r}
#tune the model with previously specified cross-validation and RMSE as target metric
tree_res <- tree_wflow %>%
                tune::tune_grid(resamples = obl_folds_spatial,
                                grid = tree_grid,
                                control = control_grid(verbose = TRUE),
                                metrics = yardstick::metric_set(rmse))

@@

#save file
DT_auto_fig_file = here("results","DT-auto-plot-OA-outlier.png")
ggsave(filename = DT_auto_fig_file, plot = DT_auto_plot)

#more detailed plot
DT_detail_plot <- tree_res %>%
                    workflowsets::collect_metrics() %>%
                    dplyr::mutate(tree_depth = factor(tree_depth)) %>%
                    ggplot2::ggplot(aes(cost_complexity, mean, color = tree_depth)) +
                             geom_line(size = 1.5, alpha = 0.6) +
                             geom_point(size = 2) +
                             facet_wrap(~ .metric, scales = "free") +
                             scale_x_log10(labels = scales::label_number()) +
                             scale_color_viridis_d(option = "plasma", begin = 0.9, end = 0) +
                             labs(x = "cost complexity",
                                  y = "mean",
                                  color = "Tree Depth")
DT_detail_plot

#save file
DT_det_fig_file = here("results","DT-det-plot-OA-outlier.png")
ggsave(filename = DT_det_fig_file, plot = DT_detail_plot)

@@

### 6. Model evaluation
```{r}
#plot the tree
rpart.plot::rpart.plot(x = workflowsets::extract_fit_parsnip(best_tree_fit)$fit,
                                                     roundint = F,
                                                     type = 5,
                                                     digits = 5,
                                                     main = "Selected Decision Tree Model",
                                                     box.palette = viridis::viridis(10, option = "D", begin = 0.9, end = 0),
                                                     shadow.col = "grey65",
                                                     col = "grey99")
#save manually as "DT-plot-OA-outlier.png" -- ggsave throws an error

#find predictions and intervals
tree_resid <- best_tree_fit %>%
                  broom.mixed::augment(new_data = obl_train_data) %>%
                  dplyr::select(.pred, log_obl_amt) %>%
                  dplyr::mutate(.resid = .pred - log_obl_amt)

#plot model predictions from tuned model versus actual outcomes
#geom_abline draws a 45 degree line, along which the results should fall
DT_pred_act <- ggplot2::ggplot(tree_resid, aes(x = .pred, y = log_obl_amt)) +
                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
                        geom_point(color = "#3b528b") +
                        scale_y_continuous(expand = c(0,0),
                                           limits = c(-30, 25)) +
                        scale_x_continuous(limits = c(0, 25)) +
                        labs(title = "Decision Tree Fit: Predicted vs. Actual Log(Obligated Funds)",
                              x = "Predicted Log(Obligated Funds) ($)",
                              y = "Actual Log(Obligated Funds) ($)") 
DT_pred_act

#save file
DT_pred_act_file = here("results","DT-pred-act-OA-outlier.png")
ggsave(filename = DT_pred_act_file, plot = DT_pred_act)

#plot model with residuals
#the geom_hline plots a straight horizontal line along which the results should fall
DT_resid <- ggplot2::ggplot(tree_resid, aes(x = as.numeric(row.names(tree_resid)), y = .resid))+
                     geom_hline(yintercept = 0, color = "red", lty = 2) +
                     geom_point(color = "#3b528b") +
                     labs(title = "Decision Tree Fit: Residuals",
                          x = "Observation Number",
                          y = "Residual")
DT_resid

#save file
DT_resid_file = here("results","DT-resid-OA-outlier.png")
ggsave(filename = DT_resid_file, plot = DT_resid)

#plot model fit vs residuals
#the geom_hline plots a straight horizontal line along which the results fall
DT_pred_resid <- ggplot2::ggplot(tree_resid, aes(x = .pred, y = .resid))+
                          geom_hline(yintercept = 0, color = "red", lty = 2) +
                          geom_point(color = "#3b528b") +
                          labs(title = "Decision Tree Fit: Residuals vs Fitted Log(Obligated Funds)",
                                x = "Predicted Log(Obligated Funds) ($)",
                                y = "Residual")
DT_pred_resid

#save file
DT_pred_resid_file = here("results","DT-pred-resid-OA-outlier.png")
ggsave(filename = DT_pred_resid_file, plot = DT_pred_resid)

@@

The identified best performing tree model has some interesting results. It identifies percent of state counties awarded IH program, total number of agencies involved, total number of counties per state, percent of counties awarded PA program, FEMA Region VI, incident month of October and March as important predictors. It is interesting that incident type or declaration type are not included as significant. The RMSE does decrease from the null, which suggests the outlier removal was a necessary step. Let's explore the other models.

@@

### 4. Tuning Using Cross-Validation and the `tune_grid()` function
```{r}
#tune the model with previously specified cross-validation and RMSE as target metric
RF_res <- RF_wflow %>%
              tune::tune_grid(resamples = obl_folds_spatial,
                              grid = RF_grid,
                              control = control_grid(verbose = TRUE, save_pred = TRUE),
                              metrics = metric_set(rmse))

@@

#save file
RF_auto_fig_file = here("results","RF-auto-plot-OA-outlier.png")
ggsave(filename = RF_auto_fig_file, plot = RF_auto_plot)
```

@@

#plot most important predictors in the model
RF_plot <- vip::vip(x_RF, 
                    num_features = 20,
                    aesthetics = list(fill = viridis::viridis(20, option = "D", begin = 0.9, end = 0)))
RF_plot

#save manually as "RF-plot-OA-outlier.png" -- ggsave throws errors

#find predictions and intervals
RF_resid <- best_RF_fit %>%
                broom.mixed::augment(new_data = obl_train_data) %>%
                dplyr::select(.pred, log_obl_amt) %>%
                dplyr::mutate(.resid = .pred - log_obl_amt)

#plot model predictions from tuned model versus actual outcomes
#geom_abline is a 45 degree line, along which the results should fall
RF_pred_act <- ggplot2::ggplot(RF_resid, aes(x = .pred, y = log_obl_amt)) +
                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) +
                        geom_point(color = "#3b528b") +
                        scale_y_continuous(expand = c(0,0),
                                           limits = c(-30, 25)) +
                        scale_x_continuous(limits = c(0, 25)) +
                        labs(title = "RF Fit: Actual vs. Predicted Log(Obligated Funds)",
                              x = "Predicted Log(Obligated Funds) ($)",
                              y = "Actual Log(Obligated Funds) ($)")
RF_pred_act

#save file
RF_pred_act_file = here("results","RF-pred-act-OA-outlier.png")
ggsave(filename = RF_pred_act_file, plot = RF_pred_act)

#plot model with residuals
#the geom_hline plots a straight horizontal line along which the results should fall
RF_resids <- ggplot2::ggplot(RF_resid, aes(x = as.numeric(row.names(RF_resid)), y = .resid))+
                            geom_hline(yintercept = 0, color = "red", lty = 2) +
                            geom_point(color = "#3b528b") +
                            labs(title = "RF Fit: Residuals",
                                  x = "Observation Number",
                                  y = "Residual")
RF_resids

#save file
RF_resid_file = here("results","RF-resid-OA-outlier.png")
ggsave(filename = RF_resid_file, plot = RF_resids)

#plot model fit vs residuals
#the geom_hline plots a straight horizontal line along which the results fall
RF_pred_resid <- ggplot2::ggplot(RF_resid, aes(x = .pred, y = .resid))+
                          geom_hline(yintercept = 0, color = "red", lty = 2) +
                          geom_point(color = "#3b528b") +
                          labs(title = "Random Forest Fit: Residuals vs Fitted Log(Obligated Funds)",
                                x = "Fitted Log(Obligated Funds) ($)",
                                y = "Residual")
RF_pred_resid

#save file
RF_pred_resid_file = here("results","RF-pred-resid-OA-outlier.png")
ggsave(filename = RF_pred_resid_file, plot = RF_pred_resid)

@@

In examining the results of the RF model within the context of RMSE, it performs marginally better than the decision tree. It identifies percent of counties awarded IH program, total number of federal agencies involved, percent of counties awarded HM program, average FEMA cost share, and whether or not the IH program was awarded to any county in the state as the five most important predictors. Let's try another model to see if we can find an improvement in the RMSE.

@@

#define the elastic net model
elastic_mod <- linear_reg(mode = "regression",
                           penalty = tune(),
                           mixture = tune()) %>%
                parsnip::set_engine("glmnet")
```

@@

### 2. Workflow Definition
```{r}
#define elastic net workflow
elastic_wflow <- workflows::workflow() %>%
                 workflows::add_model(elastic_mod) %>%
                 workflows::add_recipe(OA_rec)
```

@@

### 3. Tuning Grid Specification
```{r}
#tuning grid specification
EN_grid <- dials::grid_regular(penalty(),
                               mixture(),
                               levels = 50)
```

@@

### 4. Tuning Using Cross-Validation and the `tune_grid()` function
```{r}
#tune the model with previously specified cross-validation and RMSE as target metric
EN_res <- elastic_wflow %>%
              tune::tune_grid(resamples = obl_folds_spatial,
                              grid = EN_grid,
                              control = control_grid(verbose = TRUE, save_pred = TRUE),
                              metrics = metric_set(rmse))

#look at top 5 EN models
top_EN_models <- EN_res %>%
                    tune::show_best("rmse", n = 5)
top_EN_models

#default visualization
EN_auto_plot <- EN_res %>% autoplot()
EN_auto_plot

#save file
EN_auto_fig_file = here("results","EN-auto-plot-OA-outlier.png")
ggsave(filename = EN_auto_fig_file, plot = EN_auto_plot)
```

@@

### 5. Identify Best Model
```{r}
#select the EN model with the lowest rmse
EN_lowest_rmse <- EN_res %>%
                      tune::select_best("rmse")

#finalize the workflow by using the selected RF model
best_EN_wflow <- elastic_wflow %>%
                      tune::finalize_workflow(EN_lowest_rmse)
best_EN_wflow

#one last fit on the training data
best_EN_fit <- best_EN_wflow %>%
                    parsnip::fit(data = obl_train_data)
```

### 6. Model evaluation
```{r}
#extract model from final fit
x_EN <- best_EN_fit$fit$fit$fit

#plot most important predictors in the model
EN_plot <- vip::vip(x_EN, 
                    num_features = 20,
                    aesthetics = list(fill = viridis::viridis(20, option = "D", begin = 0.9, end = 0)))
EN_plot

#save manually as "EN-plot-OA-outlier.png" -- ggsave throws errors

#find predictions and intervals
EN_resid <- best_EN_fit %>%
                broom.mixed::augment(new_data = obl_train_data) %>%
                dplyr::select(.pred, log_obl_amt) %>%
                dplyr::mutate(.resid = .pred - log_obl_amt)

#plot model predictions from tuned model versus actual outcomes
#geom_abline is a 45 degree line, along which the results should fall
EN_pred_act <- ggplot2::ggplot(EN_resid, aes(x = .pred, y = log_obl_amt)) +
                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) +
                        geom_point(color = "#3b528b") +
                        scale_y_continuous(expand = c(0,0),
                                           limits = c(-30, 25)) +
                        scale_x_continuous(limits = c(0, 25)) +
                        labs(title = "EN Fit: Actual vs. Predicted Log(Obligated Funds)",
                              x = "Predicted Log(Obligated Funds) ($)",
                              y = "Actual Log(Obligated Funds) ($)")
EN_pred_act

#save file
EN_pred_act_file = here("results","EN-pred-act-OA-outlier.png")
ggsave(filename = EN_pred_act_file, plot = EN_pred_act)

#plot model with residuals
#the geom_hline plots a straight horizontal line along which the results should fall
EN_resids <- ggplot2::ggplot(EN_resid, aes(x = as.numeric(row.names(EN_resid)), y = .resid))+
                            geom_hline(yintercept = 0, color = "red", lty = 2) +
                            geom_point(color = "#3b528b") +
                            labs(title = "EN Fit: Residuals",
                                  x = "Observation Number",
                                  y = "Residual")
EN_resids

#save file
EN_resid_file = here("results","EN-resid-OA-outlier.png")
ggsave(filename = EN_resid_file, plot = EN_resids)

#plot model fit vs residuals
#the geom_hline plots a straight horizontal line along which the results fall
EN_pred_resid <- ggplot2::ggplot(EN_resid, aes(x = .pred, y = .resid))+
                          geom_hline(yintercept = 0, color = "red", lty = 2) +
                          geom_point(color = "#3b528b") +
                          labs(title = "EN Fit: Residuals vs Fitted Log(Obligated Funds)",
                                x = "Fitted Log(Obligated Funds) ($)",
                                y = "Residual")
EN_pred_resid

#save file
EN_pred_resid_file = here("results","EN-pred-resid-OA-outlier.png")
ggsave(filename = EN_pred_resid_file, plot = EN_pred_resid)

#print the 10 best performing hyperparameter sets
EN_res %>%
  tune::show_best(n = 10) %>%
  dplyr::select(rmse = mean, std_err)

#print the best model
EN_performance <- EN_res %>% tune::show_best(n = 1)
EN_performance

#compare model performance to null model (and other models)
EN_RMSE <- EN_res %>%
              tune::show_best(n = 1) %>%
              dplyr::transmute(
                rmse = round(mean, 3),
                SE = round(std_err, 4),
                model = "EN") %>%
             dplyr::bind_rows(RF_RMSE)
EN_RMSE
```
The Elastic Net model places great emphasis on Maine being the state receving the declaration. The next closest predictors include average FEMA cost share, Vermont, Montana, and Idaho. The EN model has an improved RMSE from the DT model but not the RF model. Let's try LASSO just to round out the analysis.

## LASSO Model
